Two-component thermal dust emission model: application to the {\it PLANCK} HFI maps

### 2014-02-04 ms.tex
clarified notation, added source function, other clean up

%the below statements seem like they might best be moved to a generic
%opening of the next section, stating our philosophy of breaking sky
%into pixels, each with 5 SED measurements, and that with this small
%number of constraints per pixel, it makes sense to have a baseline
% fit at full resolution that has only two free parameters
FDS99 found $T_2 \approx 16.2$K, $T_1 \approx 9.4$K to be typical at 
high-latitude. In the present formulation, there are only two free parameters 
to describe an observed SED: the hot dust temperature $T_2$ and the 
normalization of $M_{\nu}$. In principle, we could allow additional parameters,
 such as $f_1$ and the ratio $q_1/q_2$, to vary. However the number of free 
parameters in our baseline fit at the full $6.1'$ resolution is limited by the 
fact that, for each sky location, the observed SED contains only five 
independent intensity measurements (\verb|i100| along with the four 
highest-frequency \PLANCK~bands).

%basically just finish clarifying here that we want to start simple by fitting
%only two parameters for each pixel since we only have 5 intensity
% measurements per pixel in our baseline fits at 6.1 arcmin resoluiton

@@

\subsection{Finite Bandpass Correction}
The thermal dust emission model of $\S$\ref{sec:modeling} predicts the 
intensity $M_{\nu}$ in MJy/sr for any single frequency $\nu$. In practice, 
however, we wish to constrain our model using measurements in the broad 
\IRAS/\PLANCK~bandpasses, typically hundreds of GHz wide. We have therefore 
computed dimensionless bandpass correction factors $b_{\nu_i}$ such that 
$M_{\nu_i}b_{\nu_i}$ is the predicted observation for the \PLANCK~or \IRAS~band
 of interest, in MJy/sr. Here $\nu_i$ represents the $i^{th}$ of the five 
nominal band center frequencies for which we have broadband measurements, 
$\nu_i \in$ \{217, 353, 545, 857, 3000\} GHz. The shape of model spectrum 
$M_{\nu}$ is only dependent upon one parameter: the hot dust temperature $T_2$.
 We have used the \PLANCK~transmission curves provided in the file 
\verb|HFI_RIMO_R1.10.fits| \citep{planckresponse}. For 3000 GHz, we have used 
the \IRAS~bandpass, following the treatment of \cite{planckdust}. We note that 
it is not clear which bandpass should properly be ascribed to the \IRAS/DIRBE 
composite 100$\mu$m maps under consideration. 
% (here SFD \verb|i100|) and that of \cite{planckdust}).

@@

\subsection{Markov Chains}
As explained in $\S$\ref{sec:modeling}, the two free parameters in our 
per-pixel dust emission SED are the hot dust temperature $T_2$, which 
determinesthe shape of the SED, and the spectrum normalization. In practice,
the normalization is controlled via $\tilde{I}_{545}$, the amplitude of the 
model SED in MJy/sr at $\nu$ = 545 GHz. For each pixel, we run a 
Metropolis-Hastings Markov chain sampling the posterior probability as a 
function of the two parameters $T_2$ and $\tilde{I}_{545}$. For each pixel, we 
are sampling the posterior given by:

@@

\begin{equation}
P(\tilde{I}_{545}, T_2|\{I\}) \propto \mathcal{L}(\{I\}|\tilde{I}_{545}, T_2)P(T_2)
\end{equation}

The likelihood function is given by:
\begin{equation}
\mathcal{L}(\{I\}|\tilde{I}_{545}, T_2) = \displaystyle\prod\limits_{i}\mathcal{N}(M_{\nu_{i}}(\tilde{I}_{545}, T_2)|I_{\nu_{i}}, \sigma_{\nu_i})
\end{equation}

@@

Where the product runs over $\nu_i$ in $\{217,\ 353,\ 545,\ 857,\ 3000\}$ GHz. 
For each pixel, the errors on each intensity measurement $\sigma_{\nu_i}$ are 
given by:

@@

This error budget is the same as that of \cite{planckdust} Equation 7. 
$c_{\nu_i}$ is the multiplicative calibration uncertainty for the intensity in 
each band. We have adopted the $c_{\nu_i}$ values listed in \cite{planckdust} 
Table 1. For the 217 GHz band, we have assigned $c_{217 GHz}$ = 0.54, from 
Table 11 of \cite{planckcalib}. $CMB$ represents the CMB intensity 
subtracted at the relevant pixel. $\delta0_{\nu_i}$ represents the uncertainty
in the offset used to tie \PLANCK~intensity to HI. We have adopted the 
 $\delta0_{\nu_i}$ values of \cite{planckdust} Table 1. $n_{\nu_i}$ represents 
the instrumental noise in the pixel of interest, and is taken to be the square 
root of the \verb|ii_cov| parameter that accompanies each \PLANCK~intensity 
map. In order to obtain reasonable fitting results at high latitude where the 
data is noisy, we include a prior on $T_2$:

@@

With $\bar{T}_2$ = 16.2 K and $\sigma_{\bar{T}_2}$ = 1.4 K. In practice we 
always perform computations using the logarithm of relevant probabilities.
%mention something about the proposal distribution at some point?

@@

For each pixel, we initialize the Markov chain to have $T_2$ = $\bar{T}_2$ and
. We run 500 steps of burn-in and then 2000 steps during which we keep track
of the proposed $T_{2, j}$ and $I_{545, j}$ values. We use these 2000 values 
for each parameter to compute estimates of each parameter,  $T_2$ = 
$\langle T_{2, j} \rangle$, $I_{545}$ = $\langle I_{545, j} \rangle$, and of 
each parameter's variance, $\sigma^2_{T_2}$ = $\langle T^2_{2, j} \rangle-\langle T_{2, j} \rangle ^2$ and $\sigma^2_{I_{545}}$=$\langle I^2_{545, j} 
\rangle-\langle I_{545, j} \rangle ^2$

@@

\subsection{Calculating $\tau$}

>>>>

%the below statements seem like they might best be moved to a generic
%opening of the next section, stating our philosophy of breaking sky
%into pixels, each with 5 SED measurements, and that with this small
%number of constraints per pixel, it makes sense to have a baseline
% fit at full resolution that has only two free parameters
FDS99 found $T_2 \approx 16.2$K, $T_1 \approx 9.4$K to be typical at 
high-latitude. In the present formulation, there are only two free parameters 
to describe an observed SED: the hot dust temperature $T_2$ determines the SED 
shape and the normalization of $M_{\nu}$ determines the SED amplitude. In 
principle, we could allow additional parameters, such as $f_1$ and the ratio 
$q_1/q_2$, to vary. However the number of free parameters in our baseline fit 
at the full $6.1'$ resolution is limited by the fact that, for each sky 
location, the observed SED contains only five independent intensity 
measurements (\verb|i100| along with the four highest-frequency \PLANCK~bands).

To calculate the optical depth in the context of this model, we assume
optically thin conditions, meaning that $\tau_{\nu}$ = $M_{\nu}/S_{\nu}$, where
$M_{\nu}$ is the appropriately scaled two-component intensity and the source
function is given by:

\begin{equation}
\label{eqn:source}
S_{\nu} = \frac{f_1q_1(\nu/\nu_0)^{\beta_1}B_{\nu}(T_1) + f_2q_2(\nu/\nu_0)^{\beta_2}B_{\nu}(T_2)}{f_1q_1(\nu/\nu_0)^{\beta_1}+f_2q_2(\nu/\nu_0)^{\beta_2}}
\end{equation}
%basically just finish clarifying here that we want to start simple by fitting
%only two parameters for each pixel since we only have 5 intensity
% measurements per pixel in our baseline fits at 6.1 arcmin resolution

@@

\subsection{Finite Bandpass Correction}
The thermal dust emission model of $\S$\ref{sec:modeling} predicts the 
intensity $M_{\nu}$ in MJy/sr for any single frequency $\nu$. In practice, 
however, we wish to constrain our model using measurements in the broad 
\IRAS/\PLANCK~bandpasses, typically hundreds of GHz wide. We have therefore 
computed dimensionless bandpass correction factors $b_{\nu_i}$ such that 
$M_{\nu_i}b_{\nu_i}$ is the predicted observation for the \PLANCK~or \IRAS~band
 of interest, in MJy/sr. Here $\nu_i$ represents the $i^{th}$ of the five 
nominal band center frequencies for which we have broadband measurements, 
$\nu_i \in$ \{217, 353, 545, 857, 3000\} GHz. The shape of model spectrum 
$M_{\nu}$ is only dependent upon one parameter: the hot dust temperature $T_2$.
 We have used the \PLANCK~transmission curves provided in the file 
\verb|HFI_RIMO_R1.10.fits| \citep{planckresponse}. For 3000 GHz, we have used 
the \IRAS~bandpass, following the treatment of \cite{planckdust}. We note that 
it is not clear which bandpass should properly be ascribed to the \IRAS/DIRBE 
composite 100$\mu$m maps under consideration.

\subsection{Fitting Parameters}
In order to fit the \PLANCK/\IRAS~dust SED, we must combine the underlying 
two-component spectrum $M_{\nu}$ with the bandpass corrections $b_{\nu_i}$ to 
yield predictions which can be compared against the observed broadband 
intensities. The model-predicted broadband intensities $\tilde{I}_{\nu_i}$ are 
given by:

\begin{equation}
\label{eqn:inten}
\tilde{I}_{\nu_i} = \frac{b_{\nu_i}(T_2)M_{\nu_i}}{b_{545}(T_2)M_{545}}\tilde{I}_{545}
\end{equation}

We will treat the SED normalization $\tilde{I}_{545}$ as a free parameter in
our fitting procedure. In our baseline fits, second free parameter is $T_2$, 
which determines the shape of the SED through $M_{\nu}$ as well as 
$b_{\nu_i}(T_2)$.

% (here SFD \verb|i100|) and that of \cite{planckdust}).

@@

\subsection{Markov Chains}
As explained in $\S$\ref{sec:modeling}, the two free parameters in our 
per-pixel dust emission SED are the hot dust temperature $T_2$, which 
determines the shape of the SED, and the spectrum normalization. In practice,
the normalization is controlled via $\tilde{I}_{545}$, the amplitude of the 
model SED in MJy/sr at $\nu$ = 545 GHz. For each pixel, we run a 
Metropolis-Hastings Markov chain sampling the posterior probability as a 
function of the two parameters $T_2$ and $\tilde{I}_{545}$. For each pixel, we 
are sampling the posterior given by:

@@

\begin{equation}
\label{eqn:post}
P(\tilde{I}_{545}, T_2|\{I\}) \propto \mathcal{L}(\{I\}|\tilde{I}_{545}, T_2)P(T_2)
\end{equation}

Here \{$I$\} denotes the set of observed broadband intensities, $I_{217}$, 
$I_{353}$, etc. The likelihood function is given by:

\begin{equation}
\mathcal{L}(\{I\}|\tilde{I}_{545}, T_2) = \displaystyle\prod\limits_{i}\mathcal{N}(\tilde{I}_{\nu_{i}}|I_{\nu_{i}}, \sigma_{\nu_i})
\end{equation}

@@

Where the product runs over $\nu_i$ in $\{217,\ 353,\ 545,\ 857,\ 3000\}$ GHz
and $\tilde{I}_{\nu_i}$ is a function of $\tilde{I}_{545}$ and $T_2$ as 
specified by Equation \ref{eqn:inten}. For each pixel, the errors on each 
intensity measurement $\sigma_{\nu_i}$ are given by:

@@

This error budget is the same as that of \cite{planckdust} Equation 7. 
$c_{\nu_i}$ is the multiplicative calibration uncertainty for the intensity in 
each band. We have adopted the $c_{\nu_i}$ values listed in \cite{planckdust} 
Table 1. For the 217 GHz band, we have assigned $c_{217 GHz}$ = 0.54, from 
Table 11 of \cite{planckcalib}. $CMB$ represents the CMB intensity 
subtracted at the relevant pixel. $\delta0_{\nu_i}$ represents the uncertainty
in the offset used to tie \PLANCK~intensity to H\,\textsc{i}. We have adopted 
the  $\delta0_{\nu_i}$ values of \cite{planckdust} Table 1. $n_{\nu_i}$ 
represents the instrumental noise in the pixel of interest, and is taken to be 
the square root of the \verb|ii_cov| parameter that accompanies each 
\PLANCK~intensity map. In order to obtain reasonable fitting results at high 
latitude where the data is noisy, we include a prior on $T_2$:

@@

With $\bar{T}_2$ = 16.2 K and $\sigma_{\bar{T}_2}$ = 1.4 K. In practice we 
always perform computations using logarithms of the relevant probabilities.
%mention something about the proposal distribution at some point?

@@

For each pixel, we initialize the Markov chain with parameters $T_2$ = 
$\bar{T}_2$ and $\tilde{I}_{545}$ = $I_{545}$. We run 500 steps of burn-in and 
then 2000 steps during which we keep track of the proposed $T_{2, j}$ and 
$\tilde{I}_{545, j}$ values at the $j^{th}$ step since the end of burn-in. From
these 2000 samples, we compute estimates of each parameter, 
$T_2$ = $\langle T_{2, j} \rangle$, $\tilde{I}_{545}$ = 
$\langle \tilde{I}_{545, j} \rangle$, and of each parameter's variance, 
$\sigma^2_{T_2}$ = $\langle T^2_{2, j} \rangle-\langle T_{2, j} \rangle ^2$ and
 $\sigma^2_{\tilde{I}_{545}}$=$\langle \tilde{I}^2_{545, j} 
\rangle-\langle \tilde{I}_{545, j} \rangle ^2$. We also compute the
 corresponding estimate of the monochromatic two-component intensity at 545 
GHz, $M_{545}$ = $\langle M_{545, j} \rangle$ = 
$\langle \tilde{I}_{545, j}/b_{545}(T_{2,j}) \rangle$  and its variance. 
$M_{545}$ is of more general utility in making foreground predictions because 
it provides the normalization for the two-component spectrum $M_{\nu}$, 
independent of the \PLANCK~545 GHz bandpass. Lastly, we compute the 545 GHz
optical depth as $\tau_{545}$ = $\langle \tau_{545, j} \rangle$ = 
$\langle M_{545, j}/S_{545, j} \rangle$ and its variance, with the source 
function $S_{545}$ calculated according to Equation \ref{eqn:source}.

%in the above need to mention/introduce the use of ``j'' to denothe the j'th
%step of the MH sampling **after** burn-in has finished

% actually we I think we want to keep track of $\tilde{I}_{545}$/b_{545}
% which is the monochromatic value, since that is most useful downstream
% when making foreground predictions that typically won't have anything
% to do with Planck bandpasses

@@

%\subsection{Calculating $\tau$}
